### 2010-01-12 latex/paper1/contents.tex

%   I want less time, so I hire more programmers
%   Thee project can consume more time, due to comunication costs between
% programmers.
%   Unless there is a good division of tasks
%
% (needn't be large)

Large software systems need to be decomposed in modules in order to be
effectively maintained by groups of developers working concurrently. Software
clustering algorithms
\cite{Mancoridis1998,Anquetil1999,Tzerpos2000,Andritsos2005}, also known as
architecture recovery algorithms, find suitable decompositions by analyzing the
dependencies between components in a software system (e.g., classes in
object-oriented systems) and then grouping together highly interdependent
components.

Although many software clustering algorithms have been proposed, there is little
empirical evaluation regarding the external quality of the decompositions they
find. One important test for a software clustering algorithm is to apply it to a
collection of software systems with known reference decompositions made by
experienced developers \cite{Anquetil1999}. It is expected that a good algorithm
finds decompositions that are similar to reference decompositions, which can be
measured by a similarity metric such as MoJo \cite{Tzerpos1999} or EdgeSim
\cite{Mitchell2001}.

Unfortunately, there are few publicly available reference decompositions
\cite{Koschke2000} and, as a result, there are few studies that compare software
clustering algorithms. Also, because it is costly to obtain reference
decompositions for large systems, most studies test the algorithms with a couple
of small and medium systems \cite{Anquetil1999,Maqbool2007,Bittencourt2009}.

@@

In this paper, we present three recently proposed network models and show
empirically that they are capable of synthesizing networks that resemble
software component dependency networks. We limit our analysis to networks of
static dependencies between classes in object-oriented systems written in Java.

The paper is organized as follows: Section \ref{sec:models} presents three
network models.
%
Section \ref{sec:characterization} demonstrates a method to tell if a network
resembles a software component dependency network.
%
Section \ref{sec:evaluation} shows empirically that the three models synthesize
networks that resemble software component dependency networks.
%
Section \ref{sec:conclusion} discusses the impact of this research and future
work.

%  \begin{figure*}[!t]
%  \centering
%  \includegraphics[width=1.0\textwidth]{diagram}
%  \caption{Evaluation of a software clustering recovery algorithm}
%  \label{fig:diagram}
%  \end{figure*}

@@

Network theory studies general properties of many types of networks by using
statistical analysis. In the last decade, remarkable similarities have been
found in networks arising from domains as diverse as sociology, biology,
technology, and linguistics. It has been shown that, in many networks, the
distribution of vertex degrees is a power law, i.e., the number of vertices
connected to $k$ edges is proportional to $k^{-\gamma}$, where $\gamma$ is a
positive constant. Networks that share this property are called scale-free
networks \cite{Barabasi1999}.

Network theory has been applied to class dependency networks, represented as
directed unweighted graphs, and it was shown that they are also scale-free
\cite{Myers2003,Valverde2002}. Thus, the study of the structure of
object-oriented software systems can benefit from the vast literature available
about scale-free networks.

Scale-free network models are algorithms that can be proven, either formally or
empirically, to synthesize networks that are scale-free. In this section we
present three models that synthesize scale-free networks with built-in modular
decompositions: BCR+, CGW, and LF. The networks are represented as directed
unweighted graphs with modules, which makes them suitable for testing software
clustering algorithms.

% directed unweighted graphs

%\subsection{ER (Erdos-Renyi)}
%\subsection{Configuration model} %XXX consider using with the most typical sw

\subsection{BCR+}

The BCR model \cite{Bollobas2003} synthesizes directed scale-free networks
without modules. We have developed an extension to this model, called BCR+, that
adds modules to the construction of the network. The BCR+ model accepts the
following parameters:

\begin{itemize}
\item number of vertices, $n$;
\item a directed graph of modules, $G$;
\item three probabilities, $p_1$, $p_2$, and $p_3$, summing to 1;
\item constant $\mu$, with $0 \le \mu \le 1$;
\item base in-degree, $\din$;
\item base out-degree, $\dout$.
\end{itemize}

The graph $G$ contains one vertex for each module that will be created and
determines a ``supply'' relationship between modules. We say that module $M_2$
is a supplier of module $M_1$ if, and only if, $G$ contains an edge from $M_1$
to $M_2$.

Networks generated by the model can contain both internal edges and external
edges. An internal edge connects two vertices in the same module. An external
edge connects vertices in distinct modules. In this model, an external edge from
a vertex $v_1 \in M_1$ to a vertex $v_2 \in M_2$ is only allowed if $M_2$ is a
supplier of $M_1$.

The parameter $\mu$ controls the proportion of external edges in the network.
Lower values of $\mu$ lead to networks with fewer external edges. 
%and, therefore, to weakly coupled modules.

First the model creates a network in which each module contains exactly one
vertex. Then all allowed external edges are added. Finally, the networks is
modified by sucessive choices between three operations that add vertices or
edges to the network, until it reaches $n$ vertices. Each time, the probability
of choosing the $i$-th operation is $p_i$.

Before describing the operations, some definitions are needed. The expression
in-degree($x$) means the number of edges that enter the vertex $x$;
out-degree($x$), likewise, is the number of edges that leave vertex $x$. The
expression ``to choose a vertex according to f(x)'' means that the probability
of choosing a vertex $x$ is given by the following probability function:

$$
  \mathrm{P}(x) ~=~ \frac{ \mathrm{f}(x) }
  { \displaystyle\sum_{i} \mathrm{f}(i) }
$$

(the denominator is a normalizing factor, so the sum of probabilities is 1).

The three operations are described next:

\begin{enumerate}

\item \emph{Adding a vertex with an outgoing edge}. An existing vertex, $w$, is
chosen according to $\mathrm{f}(x) = \din + \mathrm{in\mbox{-}degree}(x)$. A new
vertex, $v$, is added to the module of $w$, together with an edge from $v$ to
$w$.

\item \emph{Adding a vertex with an ingoing edge}. An existing vertex, $w$, is
chosen according to $\mathrm{f}(x) = \dout + \mathrm{out\mbox{-}degree}(x)$. A
new vertex, $v$, is added to the module of $w$, together with an edge from $w$
to $v$.

\item \emph{Adding an edge between existing vertices}. A vertex, $v$, is choosen
from the network according to $\mathrm{f}(x) = \dout +
\mathrm{out\mbox{-}degree}(x)$.  Then an edge is added from vertex $v$ to an
existing vertex $w$, which is chosen according to one of the following cases:

\begin{enumerate}
  \item with probability $\mu$, $w$ is chosen among vertices in modules that are
  suppliers of the module of $v$;
  \item with probability $1 - \mu$, $w$ is chosen among vertices in that same
  module as $v$.
\end{enumerate}

\end{enumerate}

It is easy to see that nodes with high in-degree are more likely to receive new
ingoing edges. The parameter $\din$ can reduce this bias by providing a base
in-degree that is applied to all vertices when computing the probabilities.
Consider two vertices, $v_1$ with in-degree 4, and $v_2$ with in-degree 8. If
$\din = 0$, $v_2$ is twice more likely to receive a new incoming edge; if,
otherwise, $\din = 4$, $v_2$ is only $\frac{3}{2}$ more likely to receive the
edge. The same reasoning applies to $\dout$.

The BCR+ model is a growth model, meaning that the network is created vertex by
vertex, growing from an initial network. It can, therefore, simulate the
evolution of a software system. Moreover, it can simulate the evolution of a
software system subject to constraints in module interaction, as is the case
with top-down system design methodologies.

%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*

\subsection{CGW}

The CGW model \cite{Chen2008} was proposed to model the evolution of software
systems organized in modules. It accepts the following parameters:

\begin{itemize}
\item number of vertices, $n$;
\item number of modules, $m$;
\item probabilities $p_1, p_2, p_3, p_4$, summing to 1;
\item natural numbers $e_1, e_2, e_3, e_4$;
\item constant $\alpha$, with $\alpha \ge -1$.
\end{itemize}

Because the model is described in detail in the original paper, only an
intuitive notion of the meaning of the parameters is provided here. This model
is similar to BCR+ in which it grows an initial network by sucessive
applications of operations. In this case, there are four operations, which are
chosen according to probabilities $p_1$, $p_2$, $p_3$, and $p_4$:

\begin{enumerate}
\item adding a vertex with $e_1$ outgoing edges; 
\item adding $e_2$ edges;
\item rewiring $e_3$ edges;
\item removing $e_4$ randomly chosen edges.
\end{enumerate}

The constant $\alpha$ controls the proportion of edges that connect vertices
from distinct modules. For $\alpha = -1$, most edges will connect distinct
modules. For $\alpha > 0$, most edges will connect vertices in the same module,
and the greater the $\alpha$, the stronger the tendency. For $\alpha = 0$ there
is no bias, the two kinds of edges are equaly likely.

%Unlike BCR+, this model does not allow constraints on the connection between
%modules, however it accounts for the rewiring and the removal of edges.

%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*

\subsection{LF}

The LF model \cite{Lancichinetti2009} is a very flexible model that can
generate weighted directed networks with overlapping modules, that is, in which
a vertex can belong to more than one module. Unlike the previous models, this is
not a growth model: all vertices are generated at once and then the edges are
added.

There is also a special version of the model in which the edge weights are
discarded and the modules do not overlap. We used the original implementation of
this version, available at
\url{http://santo.fortunato.googlepages.com/inthepress2}. It accepts the
following parameters:

\begin{itemize}
\item number of nodes, $n$;
\item average in-degree, $k$, with $k < n$;
\item maximum in-degree, $max_k$, with $k \le max_k < n$;
\item mixing parameter, $\mu$, with $0 \le \mu \le 1$;
\item minus exponent for the degree distribution, $\gamma$;
\item minus exponent for the module size distribution, $\beta$;
\item minimum module size, $min_m$;
\item maximum module size, $max_m$, with $max_m \ge min_m$;
\end{itemize}

The sizes of the modules are selected from a power law distribution with
exponent $-\beta$. The mixing parameter, $\mu$, is the proportion of edges in the
network that connect vertices from distinct modules. In this model, some
combinations of parameter values are unfeasible. For example, if $n = 100$ then
$min_m$ cannot be 60 (otherwise there would be modules smaller than the minimum
module size).

@@

In this section we present a similarity metric and validate it by appplying it
to a data set containing both software and non-software networks.

@@

In a recent work, Milo et al. \cite{Milo2002} proposed to characterize networks
by analyzing their triad concentration. A triad is a network with three vertices
in which all vertices are connected. There are only 13 distinct triads, one for
each possible configuration of directed edges, as shown in Figure
\ref{fig:triads}.

@@

% Word adjacency network: word Y follows word X => X->Y
\begin{figure}[t]
\center
\includegraphics{tcp}
\caption{Triad concentration profiles (TCP) for two networks. On the left,
network extracted from the software system JabRef (see the Appendix). On the
right, word adjacency network for the Japanese language \cite{Milo2004}.}
\label{fig:profiles}
\end{figure}

Following another work by Milo et al. \cite{Milo2004}, similarity between two
networks can be measured by computing Pearson's correlation coefficient between
the corresponding TCPs, which yields a value between -1 (most dissimilar) and 1
(most similar):

@@

To support the evaluation of the metric, we have collected 131 networks from
many different domains. The networks are described in detail in the Appendix.

\subsubsection{Software networks} We have collected 65 software systems written
in Java, with size ranging from 111 to 35,363 classes. Java was chosen for being
a popular programming language in which many open source systems have been
written. The software networks, representing static dependencies between
classes, were extracted with the tool Dependency Finder\footnote{Available at
\url{http://depfind.sf.net}.}.

\subsubsection{Non-software networks} We have collected 66 networks from
distinct domains, such as biology, sociology, technology, and linguistics, with
size ranging from 32 to 18163 vertices. These networks are freely available on
the Internet and have been previously studied in the literature.
% S-score

@@

For the purposes of this research, a similarity metric must fullfil two
conditions: (i) it must yield high similarity between software networks, and
(ii) it must yield lower similarity between software networks and networks from
other domains.

Using the data set we can define S-score, a metric that represents how much a
particular network resemble software networks. It is defined as the average
similarity between the network and a sample of software networks:

@@

We used the tool igraph \cite{igraph} to extract the TCP for each network in the
data set. Then, we measured the S-score for each software network, which ranged
from 0.83 to 0.98. The average S-score was 0.97 and the standard deviation,
0.03. The high average S-score and the low standard deviation show that the
metric successfully characterizes software networks by capturing their common
structural patterns.

%TODO: Boxplot by network class (software, metabolic, neural, social etc.)

@@

\subsection{A Network Classification Model} \label{sec:classmodel}

Although the S-score of a network tells how close it is from software networks,
it does not tell whether a network is close enough that it can be considered
software-like. What is needed is a binary classification model that
distinguishes software-like networks from the other networks. The distinction
can be made by choosing a suitable S-score threshold. Networks with S-score
below the threshold are considered dissimilar from software networks; only
networks with S-score above the threshold are considered software-like.

@@

To get 100\% precision, the threshold needs to be 0.98, so the non-software
network with highest S-score is below the threshold. The recall in this case,
though, would be too low, because most software networks would be misclassified.
So we chose the value 0.88, that is immediatly above the second greater S-score
for a non-software network. With this value, we have both high recall (95.4\%)
and high precision (96.9\%).

%threshold 0.88: 2/66 non-sw misclassified, 3/65 sw misclassified
%precision: 62 / 64
%recall: 62 / 65

@@

\section{Evaluation of Network Models} \label{sec:evaluation}
% v.  differentiate, differ; distinguish; contrast; discern

In the previous section it was shown that many networks, although scale-free,
can be distinguished from software networks by a simple classification model
based on triad concentration profiles. In this section we show empirically that
the three network models previously presented can synthesize networks that are
indistinguishable from software networks. The experiment consists of
synthesizing networks using many combinations of parameters from the three
models, and then classifying each network as software-like or non software-like.
%After that, we present rules for the choice of parameter values that lead to
%software-like networks.  \subsection{Synthetic Data Set}
%
%We want to investigate if, with a proper choice of parameters, a model is
%capable of synthesizing a network that resembles software networks. 
Because the possible combinations of parameter values are infinite, we have set
the number of vertices to 1000 and then varied the remaining parameters in
discrete steps.
%In this section we describe the combinations of parameters values used for each
%model.

\subsection{BCR+ networks}

We have chosen five different module graphs, which where extracted from actual
dependencies between Java archives distributed with five different software
systems of our sample: GEF (2 archives), iBATIS (4 archives), MegaMek (8
archives), findbugs (16 archives), and zk (32 archives). Because many of these
archives are intended to be reused in distinct projects, they provide a good
coarse-grained approximation to the concept of module.

%The module dependency networks are shown on Figure \ref{fig:architectures}.
%\begin{figure*}[!t] \centering
%\includegraphics[width=1.0\textwidth]{architectures} \caption{Module dependency
%graphs} \label{fig:architectures} \end{figure*}

For the remaining parameters, the following values were chosen:

\begin{itemize}
\item $p_1, p_2, p_3 \in \{0.0, 0.2, 0.4, 0.6, 0.8, 1.0\}$, with $p + q + r = 1$ and
$p + q > 0$;
\item $\din, \dout \in \{0, 1, 2, 3, 4\}$;
\item $\mu \in \{0.0, 0.2, 0.4, 0.6\}$.
\end{itemize}

We chose $p + q > 0$ because otherwise no node would be added to the initial
network. We also avoided large values for $\mu$ in order to ignore networks with
strongly coupled modules.

In total, 9,500 BCR+ networks were synthesized.

\subsection{CGW networks}

The parameters values were chosen as such:

\begin{itemize}
\item $p1, p2, p3, p4 \in \{0.0, 0.2, 0.4, 0.6, 0.8, 1.0\}$, with $p1 > 0$ and
$p1 + p2 + p3 + p4 = 1$;
\item $e1, e2, e3, e4 \in \{1, 2, 4, 8\}$;
\item $\alpha \in \{-1, 0, 1, 10, 100, 1000\}$;
\item number of modules: $m \in \{2, 4, 8, 16, 32\}$.
\end{itemize}

Note that $e_i$ has no effect when $p_i = 0$; in this case $e_i$ was just set to
zero.

In total, 38,790 CGW networks were synthesized.

\subsection{LF networks}

The following values were chosen:

\begin{itemize}
  \item mixing parameter: $\mu \in \{0.0, 0.2, 0.4, 0.6\}$;
  \item degree exponent distribution: $\gamma \in \{2.18, 2.70, 3.35\}$;
  \item module size distribution: $\beta \in \{0.76, 0.99, 1.58\}$;
  \item average degree: $k \in \{5, 10, 15, 25\}$;
  \item maximum degree: $max_k \in \{58, 157, 482\}$;
  \item minimum module size: $min_m \in \{1, 10, 273\}$.
\end{itemize}

In order to choose these values, we analyzed software networks from our sample
with approximately 500 to 2,000 nodes, so no network was much bigger or much
smaller than the synthetic networks. We computed the exponents for the degree
and module size distributions using the maximum likelihood estimation method
\cite{Clauset2007}, and then chose the minimum, median and maximum values. For
$k$, $maxk$, and $min_m$, the values extracted from the software networks were
divided by the number of nodes in the network and then multiplied by 1000. We
then selected the minimum, median and maximum values. The parameter $max_m$ was
left unbound to avoid impossible combinations of parameters.

In total, 1,296 LF networks where synthesized.

\subsection{Results}
% This subsection needn't to be big. It's just numbers.

Each synthesized network was classified as software-like or non software-like,
using the classification model presented in section \ref{sec:classmodel}. The
results are summarized in Table \ref{tab:results}.

\begin{table}
\caption{Results for the classification of synthetic networks}
\centering
\begin{tabular}{|l|l|}
\hline
Model & Networks classified as software-like \\
\hline 
\hline
BCR+ & 21.18\% \\ % 2012 / 9500
\hline
CGW  & 19.40\% \\  % 7524 / 38790
\hline
LF   & 31.25\% \\ %  405 / 1296
\hline
\end{tabular}
\label{tab:results}
\end{table}

All models synthesized both software-like and non software-like networks. The
proportion of software-like networks was greater than 19\% for all models,
discarding the possibility that this result was obtained by pure chance.

%Show some graphs: histogram of average correlations for each model.
%\ref{fig:histograms}
%\begin{figure*}[!t]
%\center
%
%\subfigure[CGW]{\includegraphics[width=1.0in]{hist-cgw}}
%
%\subfigure[BCR+]{\includegraphics[width=1.0in]{hist-bcr}}
%
%\subfigure[LF]{\includegraphics[width=1.0in]{hist-lf}}
%
%\caption{Triad profiles}
%\label{fig:histograms}
%\end{figure*}

@@

The use of synthetic data is common in distributed computing research, but still
underexplored in software engineering research. Because many reverse engineering
tasks rely on dependency data \cite{Tonella2007}, we expect this work to have
impact beyond the software clustering community.

%Although the use of models and synthetic data is somewhat common in distributed
%computing research, it is underexplored in software engineering research. We
%expect this work to contribute to explore t
%common in distributed computing research, simulation is underexplored
%in software engineering. This work opens a door to the usage of simulation in
%the field of reverse engineering of software in order to evaluate RE 
%algorithms.

@@

\appendix %{Network Data Set}

This appendix lists the networks used in this work.

\subsection{Software Networks}

Systems hosted by SourceForge (\url{http://sourceforge.net/}):
\begin{itemize}
\item AbaGuiBuilder-1.8
\item alfresco-labs-deployment-3Stable
\item aoi272
\item stendhal-0.74
\item battlefieldjava-0.1
\item checkstyle-5.0
\item dom4j-1.6.1
\item findbugs-1.3.8
\item freetts-1.2.2-bin
\item ganttproject-2.0.9
\item geoserver-2.0-beta1-bin
\item geotools-2.5.5-bin
\item gfp\_0.8.1
\item hibernate-distribution-3.3.1.GA-dist
\item hsqldb\_1\_8\_0\_10
\item iBATIS\_DBL-2.1.5.582
\item iReport-nb-3.5.1
\item JabRef-2.5b2-src
\item jailer\_2.9.9
\item jalopy-1.5rc3
\item jasperreports-3.5.2-project
\item jfreechart-1.0.13
\item pentaho-reporting-engine-classic-0.8.9.11
\item jGnash-2.2.0
\item jgraphpad-5.10.0.2
\item jmsn-0.9.9b2
\item juel-2.1.2
\item JXv3.2rc2deploy
\item makagiga-3.4
\item MegaMek-v0.34.3
\item iFreeBudget-2.0.9
\item mondrian-3.1.1.12687
\item oddjob-0.26.0
\item openxava-3.1.2
\item pdfsam-1.1.3-out
\item pjirc\_2\_2\_1\_bin
\item pmd-bin-4.2.5
\item proguard4.3
\item smc\_6\_0\_0
\item squirrel-sql-3.0.1-base
\item squirrel-sql-3.0.1-standard
\item tvbrowser-2.7.3-bin
\item villonanny-2.3.0.b02.bin
\item rapidminer-4.4-community
\item zk-bin-3.6.1
\end{itemize}

Systems hosted in other sites:

\begin{itemize}
\item ArgoUML-0.28
\item GEF-0.13-bin
\item Hl7Comm.1.0.1
\item IRPF2009v1.1
\item broker-4.1.5 (OurGrid)
\item dbwrench
\item ec2-api-tools
\item ermodeller-1.9.2-binary
\item flyingsaucer-R8
\item gdata-src.java-1.31.1
\item guice-2.0
\item gwt-windows-1.6.4
\item jai-1\_1\_4-pre-dr-b03-lib-linux-i586-08\_Jun\_2009
\item jakarta-tomcat-5.0.28-embed
\item juxy-0.8
\item myjgui\_0.6.6
\item peer-4.1.5 (OurGrid)
\item subethasmtp-3.1
\item thinkui\_sqlclient-1.1.2
\item worker-4.1.5 (OurGrid)
\end{itemize}

\subsection{Networks from Other Domains}

\begin{itemize}
\item 5 friendship networks from Facebook \cite{Traud2008}
\item 3 electronic circuit networks \cite{Milo2004}
\item 4 word adjacency networks \cite{Milo2004}
\item 3 protein structure networks \cite{Milo2004}
\item 2 social networks of positive sentiment \cite{Milo2004}
\item 43 metabolic networks \cite{Jeong2000}
\item Protein interaction network for yeast \cite{Jeong2001}
\item Links between political blogs \cite{Adamic2005}
\item Neural network of C Elegans \cite{Watts1998}
\item Network ``beta3sreduced'' (unknown source)
\item Network ``czech'' (unknown source)
\item Network ``ecoli-metabolic'' (unknown source).
\end{itemize}

>>>>

Large software systems need to be decomposed in modules in order to be
effectively maintained by groups of developers working concurrently. Software
clustering algorithms \cite{Anquetil1999}, also known as architecture recovery
algorithms, find suitable decompositions by analyzing the dependencies between
components in a software system (e.g., classes in object-oriented systems) and
then grouping together highly interdependent components.

Although many software clustering algorithms have been proposed, there is
little empirical evaluation regarding the external quality of the
decompositions they find, i.e., how close the decompositions are from reference
decompositions made by experienced developers \cite{Anquetil1999}. It is
expected that a good algorithm finds decompositions that are similar to
reference decompositions, which can be measured by a similarity metric such as
MoJo or EdgeSim \cite{Mitchell2001}.

Unfortunately, there are few publicly available reference decompositions
\cite{Anquetil1999} and, as a result, there are few studies that compare
software clustering algorithms. Also, because it is costly to obtain reference
decompositions for large systems, most studies test the algorithms with a
couple of small and medium systems \cite{Anquetil1999,Bittencourt2009}.

@@

In this paper, we show that three recently proposed network models are capable
of synthesizing networks that resemble software component dependency networks.
For the sake of simplicity, we limit our analysis to networks of static
dependencies between classes in object-oriented systems written in Java.

The paper is organized as follows: Section \ref{sec:models} summarizes main
characteristics of the three network models.
%
Section \ref{sec:characterization} demonstrates a method to tell if a network
resembles a software component dependency network.
%
Section \ref{sec:evaluation} shows empirically that the three models synthesize
networks that resemble software component dependency networks.
%
Section \ref{sec:conclusion} discusses the impact of this research and future
work.

@@

Network theory studies general properties of many types of networks by using
statistical analysis. Of special interest are the so called scale-free
networks, i.e., networks with a highly heterogeneous distribution of the number
of edges per vertex. Scale-free networks have been found in many domains,
including sociology, biology, technology, linguistics, and software. It has
been found that class dependency networks, represented as directed unweighted
graphs, are scale-free \cite{Myers2003}. Thus, the study of the structure of
object-oriented software systems can benefit from the vast literature available
about scale-free networks.

Scale-free network models are algorithms that can be proven, either formally or
empirically, to synthesize networks that are scale-free. Although many
scale-free network models have been proposed, we were able to find in the
literature only two models that generate networks with built-in modules: CGW
\cite{Chen2008} and LFR \cite{Lancichinetti2009}. Also, we adapted a simple and
flexible model, BCR \cite{Bollobas2003}, to produce modules. We called the new
model BCR+.

The networks generated by the three models are represented as directed
unweighted graphs with modules, which makes them suitable for testing software
clustering algorithms. All models are controllable by a set of parameters,
including number of vertices and number of modules. Parameters specific to each
model influence network characteristics such as number of edges and proportion
of external edges (edges connecting vertices from distinct modules).

@@

In this section we present a similarity metric and validate it by applying it
to a data set containing both software and non-software networks.

@@

In a recent work, Milo et al. \cite{Milo2004} characterized networks by
analyzing their triad concentration. A triad is a network with three vertices
in which all vertices are connected. There are only 13 distinct triads, one for
each possible configuration of directed edges, as shown in Figure
\ref{fig:triads}.

@@

% Word adjacency network: word Y follows word X => X->Y
\begin{figure}[t]
\center
\includegraphics{tcp}
\caption{Triad concentration profiles (TCP) for two networks. On the left,
network extracted from the software system JabRef, version 2.5b2. On the
right, word adjacency network for the Japanese language \cite{Milo2004}.}
\label{fig:profiles}
\end{figure}

Following the work by Milo et al. \cite{Milo2004}, similarity between two
networks is measured by computing Pearson's correlation coefficient between the
corresponding TCPs, which yields a value between -1 (most dissimilar) and 1
(most similar):

@@

To support the evaluation of the metric, we have collected 131 networks from
many different domains. The networks are described in detail at
\url{http://bit.ly/7DPW5X}.

\subsubsection{Software networks} We have collected 65 software systems written
in Java, with size ranging from 111 to 35,363 classes. Java was chosen for
being a popular programming language in which many open source systems have
been written. The software networks, representing static dependencies between
classes, were extracted with the tool Dependency Finder
(\url{http://depfind.sf.net}).

\subsubsection{Non-software networks} We have collected 66 networks from
distinct domains, such as biology, sociology, technology, and linguistics, with
size ranging from 32 to 18,163 vertices. These networks are freely available on
the Internet and have been previously studied in the literature.
% S-score

@@

For the purposes of this research, a similarity metric must fulfill two
conditions: (i) it must yield high similarity between software networks, and
(ii) it must yield lower similarity between software networks and networks from
other domains.

Using the data set we can define S-score, a metric that represents how much a
particular network resembles software networks. It is defined as the average
similarity between the network and a sample of software networks:

@@

We used the tool igraph (\url{http://igraph.sf.net/}) to extract the TCP for
each network in the data set. Then, we measured the S-score for each software
network, which ranged from 0.83 to 0.98, with average 0.97 and standard
deviation 0.03. The high average S-score and the low standard deviation show
that the metric successfully characterizes software networks by capturing their
common structural patterns.

@@

\subsection{A Network Classification Model} \label{sec:classmodel} % TODO:
Although the S-score of a network tells how close it is from software networks,
it does not tell whether a network is close enough that it can be considered
software-like. What is needed is a binary classification model that
distinguishes software-like networks from the other networks. The distinction
can be made by choosing a suitable S-score threshold. If a network has a S-score
below the threshold, it is considered dissimilar from software networks;
otherwise, it is considered software-like.

@@

To get 100\% precision, the threshold needs to be 0.98, so the non-software
network with highest S-score is below the threshold. But the recall in this
case would be low, because most software networks would be misclassified, so we
chose the value 0.88, that is immediately above the second greater S-score for
a non-software network. With this value, we have both high recall (95.4\%) and
high precision (96.9\%).

@@

\section{Evaluation of Network Models} \label{sec:evaluation}

In the previous section it was shown that many networks, although scale-free,
can be distinguished from software networks by a simple classification model
based on triad concentration profiles. In this section we show empirically that
the three network models previously presented can synthesize networks that are
indistinguishable from software networks. The experiment consists of
synthesizing networks using many combinations of parameters from the three
models, and then classifying each network as software-like or non
software-like.

Because the possible combinations of parameter values are infinite, we have set
the number of vertices to 1000 and then varied the remaining parameters in
discrete steps. The number of modules varied between 2, 4, 8, 16, and 32. For
each of the remaining model-specific parameters, at least 3 values were chosen.
When possible, the values were chosen to cover the entire parameter domain. In
the case of unbound parameters, the values were chosen to approximate
characteristics of the networks in the software network data set. In total,
9,500 networks were generated with the BCR+ model, 38,790 with the CGW model,
and 1,296 with the LFR model.

\subsection{Results}

Each synthesized network was classified as software-like or non software-like,
using the classification model presented in Section \ref{sec:classmodel}. The
results are summarized in Table \ref{tab:results}.

\begin{table}
\caption{Results for the classification of synthetic networks}
\centering
\begin{tabular}{|l|l|}
\hline
Model & Networks classified as software-like \\
\hline 
\hline
BCR+ & 21.18\% \\ % 2012 / 9500
\hline
CGW  & 19.40\% \\  % 7524 / 38790
\hline
LFR  & 31.25\% \\ %  405 / 1296
\hline
\end{tabular}
\label{tab:results}
\end{table}

All models synthesized both software-like and non software-like networks. The
proportion of software-like networks was greater than 19\% for all models,
discarding the possibility that this result was obtained by pure chance. (The
specific proportion of software-like networks for each network should not be
interpreted as a measure of quality: with these results we cannot tell whether
one model is better than the others.)

@@

The use of synthetic data is common in distributed computing research, but
still underexplored in software engineering research. Because many reverse
engineering tasks rely on dependency data, we expect this
work to have impact beyond the software clustering community.

@@

% Also, it is worth investigating models that account for more than one type
% of dependency.
